---
title: "Advanced options"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview 

The topics in this document are:

* [Data management 2](#datamanagement2)

* [Advanced graphs](#graphs2)

## Data management 2 {#datamanagement2}
We can do a lot of calculations and data manipulations with R. 
First of all, we can use R as a calculator:

```{r echo = T, eval=T} 
2 + 4
``` 
 
We can save the result as an *object* in R. Then it is displayed in the upper right pane.
With `<-` we assign an object

```{r echo = T, eval=T} 
x <- 2 + 4
x
``` 
&nbsp;


##### Vectors
Vectors are a series of different values. With `c()` we can combine elements into a vector. Let's say I want to make a vector "v" with 4 values; 1,2,3,4. 

```{r echo = T, eval=T} 
v <- c(1,2,3,4)
v
class(v)
```
Since the elements of "v" are numbers, the class is numeric. I can also make a vector of text elements. In that case, the class of the vector will be "character".

```{r echo = T, eval=T} 
v2 <- c("Monday", "Tuesday", "Wednesday", "Friday")
v2
class(v2)
```

The function `c()` can also be used to combine two vectors.

```{r echo = T, eval=T} 
v3 <- c(v,v2)
v3
class(v3)
```

Notice how all the elements of v3 are now strings. 

&nbsp;
&nbsp;

##### Matrices and data frames
A matrix is a two dimensional vector. We can make a matrix with the function `matrix()`. 

```{r echo = T, eval=T} 
m1 <- matrix(c(1:9), nrow = 3)
m1
class(m1)
```
The matrix `m1` is filled with the number 1 through 9 (with `c(1:9)` I make a sequence of numbers). The matrix is filled in vertically. By specifying `nrow = 3` R knows how many rows I want. Now compare this with the following matrix:

```{r echo = T, eval=T} 
m2 <- matrix(c(1:9), nrow = 3, byrow = TRUE)
m2
```

Can you guess what `byrow = TRUE` does?

&nbsp;

For our data sets, we often use the class `data.frame`. I can make my matrix into a data frame:

```{r echo = T, eval=T} 
data2 <- as.data.frame(m2)
data2
class(data2)
```

You can also build your data set by combining vectors using the `cbind()` function:

```{r echo = T, eval=T} 
data3 <- as.data.frame(cbind(v, v2))
data3
```

**Note** Notice the difference between `data3` and the vector `v3` we created earlier. 



Certain row numbers or colum numbers can be accessed using the square brackets `[,]`. 
A number before the comma gives this row number. This code gives me the first  
```{r echo = T, eval=T} 
 mtcars[1,] 
``` 
 
 
## Advanced graphs {#graphs2}
 
 Coming soon!