---
title: "Getting started"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

For a lot of R functions, additional libraries need to be loaded.

This can be done with:

```{r echo = T, eval=FALSE}
library(lattice)
```

Sometimes the library needs to be installed first:

```{r echo = T, eval=FALSE}
install.packages(lattice)
library(lattice)
```

*Installing* a library only needs to be done once. *Loading* the library must be done every time you start R. 



# Importing data sets

How do I get my data in R?
### SPSS file

1) Load the library foreign
```{r echo = T, eval=FALSE}
library(foreign)
```

2) Load data from spss using the function read.spss
```{r echo = T, eval=FALSE}
data <- read.spss("C:/Documents/data.sav", to.data.frame = TRUE)
```

### Excel file

1) Load the library readxl
```{r echo = T, eval=FALSE}
library(readxl)
```

2) Load data from spss using the function read_excel
```{r echo = T, eval=FALSE}
data <- read_excel("C:/Documents/data.xlsx")
```


**Note 1** For R to recognize the folder you need to use forward slashes (`/`) instead of backward slashes (`\`)

**Note 2** Pay attention to the name you give your data set. You want it to be distinct, but don't make it too long!

## Importing data with the menu
Data can also be loaded via the menu. Go to File -> Import Dataset -> choose your filetype.



# Graphs
Let's explore a dataset graphically.

We will use a standard data set available in R (mtcars).
```{r echo = T, eval=T}
data(mtcars)
```

With the function `head`, we can print the first couple of rows; check out the names of the variables and the first entries.

```{r echo = T, eval=T}
head(mtcars)
```

Make a simple histogram (mpg = continuous variable)

```{r echo = T, eval=T}
hist(mtcars$mpg)
```

Make a simple scatter plot (mpg & disp = continuous variables)
```{r echo = T, eval=T}
plot(mtcars$mpg ~ mtcars$disp)
```

Make a simple boxplot (mpg = continuous variable, cyl = categorical variable)
```{r echo = T, eval=T}
boxplot(mtcars$mpg ~ mtcars$cyl)
```


Improve the scatterplot by changing the color, adding a title, etc. 

```{r echo = T, eval=T}
plot(mtcars$mpg ~ mtcars$disp, xlab = "Disp", ylab = "Mpg", 
     xlim = c(75, 475), ylim = c(10, 35),
     main = "Scatterplot Mpg - Disp", col = "red", pch = 20, cex = 1.5)
```